---
title: "Efficiently Cleaning and Labelling Data Columns with `thekidsbiostats`"
author: "Zac Dempsey, Dr Robin van de Meeburg"
format:
  html:
    code-fold: true
    tabset: true
    toc: true
    toc-location: left
date: "2026-02-12"
categories:
  - R
  - Data cleaning
  - Package
draft: true
css: _extensions/html/styles.css
editor_options: 
  chunk_output_type: inline
execute:
  echo: false
  warning: false 
fig-align: center
fig-format: png
---

```{r echo=TRUE}
#| code-fold: false
library(thekidsbiostats) # install with install.packages("thekidsbiostats", repos = "the-kids-biostats.r-universe.dev")
```

# Overview

Datasets rarely come with clean and consistent formatting in their column names. Therefore, it is common practice when performing analysis to rename the original column names to syntactic, code-friendly names (i.e., unique, lower-case names with no spaces or special characters). Alongside these code-friendly columns, it is then also often useful to store *label attributes*, which provide clean, human-readable versions labels to for each code-friendly variable. These can be called upon later for end-user output, such as analytical reports, tables and figures. 

While a variety of functions for **renaming columns** (e.g., `dplyr::rename()`) and **applying labels** (e.g., using the [`labelled`](https://cran.r-project.org/web/packages/labelled/index.html) or [`Hmisc`](https://cran.r-project.org/web/packages/Hmisc/index.html) packages) exist, a reasonable amount of leg work is still needed before such functions can be applied.

This post will demonstrate how [`thekidsbiostats`](https://github.com/The-Kids-Biostats/thekidsbiostats) can be used to streamline and expedite the process of standardising and labelling columns. Specifically, the following functions will be introduced:

+ [`make_column_dict()`](https://the-kids-biostats.github.io/thekidsbiostats/reference/make_column_dict.html)–--create a copy-pasteable column name dictionary template.
+ [`update_columns()`](https://the-kids-biostats.github.io/thekidsbiostats/reference/update_columns.html)–--rename and label columns using a column dictionary.

# Motivating Example

As an example, let's point to the `data_patient` dataset, an example housed within the package:

```{r echo=TRUE}
#| code-fold: false
data(data_patient, package = "thekidsbiostats")
```

Previewing this dataset:

```{r}
data_patient %>%
  thekids_table()
```

Immediately, we see that the column headers are not formatted in a very nice way, and especially not for programming! The column headers contain (any combination of):

+ Categorical variable coding,
+ Full questions,
+ Spacing,
+ Inconsistent letter capitalisation,
+ Punctuation (parentheses, equals signs, questions marks, dashes, slashes).

The spaces and punctuation in these column names require us to use backticks (``) to avoid R interpreting these symbols literally. 

Beyond visually looking messy, it can also be time consuming and tricky to reference these column names when attempting to use them in operations. For example, using the original column names, let's implement two trivial manipulations/changes to the data:

::: panel-tabset
## i) Calculate BMI

```{r echo=TRUE}
#| code-fold: false
data_patient$bmi <- data_patient$`WHAT IS YOUR CURRENT WEIGHT? (kg)` / (data_patient$`WHAT IS YOUR HEIGHT? (cm)` / 100.0)^2
```

## ii) Re-Code `Sex`

Recode the Sex column by replacing 0/1 with Male/Female, as defined in the existing column header.

```{r echo=TRUE}
#| code-fold: false
data_patient$`Sex (0=Male, 1=Female)` = case_when(
  data_patient$`Sex (0=Male, 1=Female)` == 0 ~ "Male",
  data_patient$`Sex (0=Male, 1=Female)` == 1 ~ "Female",
  TRUE ~ NA_character_
)
```
:::
::: column-margin
Clearly, coding each of these changes are cumbersome and time intensive, and can restrict the readability of our code.
:::


Functions such as `janitor::clean_names()` can be very useful for standardising such column names into syntactic names. However, this approach is still imperfect, and the resulting names may remain just as problematic (as seen below).

```{r echo=TRUE}
#| code-fold: false
colnames(janitor::clean_names(data_patient[6]))
```
::: column-margin
The "cleaned" version of this column name simply omits spaces and punctuation, and is still far too long.
:::


## Step 1: Create Data Dictionary

Ideally, we want to (i) rename the columns, and (ii) simultaneously retain human-readable labels for presentation/reporting.

Therefore, our solution is to manually set up a dictionary that maps the **old column names** to **newly formatted column names** as well as **apply label attributes**.

The `make_column_dict()` function facilitates this process by creating a ready-to-fill template based on the original column names in the data. 

For example:

```{r echo=TRUE}
#| code-fold: false
dict <- make_column_dict(data = data_patient,
                         auto_clean = TRUE)
```
::: column-margin
The output of `make_column_dict` is a `tribble`---a *row-wise* data-frame object which allows one to more easily visualise the format of the table.
:::

We can either directly add to, or modify, the `dict` object (as assigned above) or copy-paste the output into our code and modify the new column names and labels as needed (as done below):

```{r echo=TRUE}
#| code-fold: false
dict_patient <- tribble(
  ~old, ~new, ~label,
  'Patient ID#', 'id', 'Patient ID',
  'DOB [YYYY-MM-DD]', 'dob', 'Date of Birth',
  'Sex (0=Male, 1=Female)', 'sex', 'Sex',
  'WHAT IS YOUR HEIGHT? (cm)', 'height_cm', 'Height (cm)',
  'WHAT IS YOUR CURRENT WEIGHT? (kg)', 'weight_kg', 'Weight (kg)',
  'Do you currently smoke any form of tobacco products, including cigarettes, cigars, or pipes, on a regular basis?', 'current_smoking', 'Current tobacco consumption',
  'bp (mmHg)', 'bp_mmhg', 'BP (mmHg)',
  'Cholesterol / mmolL', 'cholesterol_mmol_l', 'Cholesterol (mmolL)',
  'bmi', 'bmi', "BMI (kg/m^2)"
)
```
::: column-margin
This `tribble` has been manually modified with "clean" column names and variable labels. Additionally, a row for the derived `bmi` variable has been included.
:::

::: callout-important
Naively renaming columns for easy manipulation in code risks dropping potentially useful information contained in the original column name (e.g., units and symbols). It is important to have a *data dictionary* to retain this information about variables, which is distinct from the column name dictionary we're creating here. For example, neither our `new` or `label` value for the `sex` variable above captures the coding.  
:::

Alternatively, we can also write this file (e.g., csv, rds, xlsx) to our machine for editing, by specifying `file`:

```{r echo=TRUE, eval=FALSE}
#| code-fold: false
make_column_dict(data = data_patient,
                 auto_clean = TRUE,
                 file = "colnames.csv")
```

And then read the csv back in after new column names and labels have been added:

```{r echo=TRUE}
#| code-fold: false
dict_patient <- readr::read_csv(file = "colnames.csv")
```

```{r}
dict_patient %>%
  thekids_table()
```

## Step 2: Apply column dictionary

Now that we have defined the column dictionary above, we can simply use `update_columns()` to apply these changes in a single line.

```{r echo=TRUE}
#| code-fold: false
data_clean <- update_columns(
  data_patient, 
  dict = dict_patient,
  old = "old",    # Name for *old* column names in `dict_patient` (default)
  new = "new",    # Name for *new* column names in `dict_patient` (default)
  label = "label" # Label attribute for each variable (default)
  )
```
::: column-margin
The arguments `old`, `new`, `label` must correspond to the column names of the data dictionary object. By default, the names assigned in `make_column_dict()` are assumed. If the data dictionary uses different header names, these must be explicitly specified in `update_columns()`.
:::

## Previewing Output

Previewing this "cleaned" dataset:

```{r}
data_clean %>%
  labelled::remove_labels() %>%
  thekids_table()
```

Because the labels we specified are stored as attributes, they are automatically recognised by many summary packages that respect label metadata (e.g., `gtsummary::tbl_summary()` tables).

```{r echo=TRUE}
#| code-fold: false
data_clean %>%
  tbl_summary(include = -id,
              type = list(height_cm ~ "continuous",
                          weight_kg ~ "continuous",
                          cholesterol_mmoll ~ "continuous",
                          bmi ~ "continuous")) %>%
  thekids_table()
```


# Closing comments

Overall, the `make_column_dict()` and `update_columns()` functions from our package offer an efficient solution for transforming and labelling datasets with problematic column headers. More concise column names are easier to work with and enhance the readability/tractability of code, while user-defined labels allow users to easily produce more polished and publication-ready outputs.

# Acknowledgements

Thanks to Dr Matt Cooper, Wesley Billingham, Dr Robin Cook and Dr Elizabeth McKinnon for providing feedback on and reviewing this post.

## Reproducibility Information

To access the .qmd (Quarto markdown) files as well as any R scripts or data that was used in this post, please visit our GitHub:

<https://github.com/The-Kids-Biostats/The-Kids-Biostats.github.io/tree/main/posts/>

The session information can also be seen below.

```{r echo = T}
sessionInfo()
```
