---
title: "Efficiently Cleaning and Labelling Columns with `thekidsbiostats`"
author: "Zac Dempsey, Dr Robin van de Meeburg"
format:
  html:
    code-fold: true
    tabset: true
    toc: true
    toc-location: left
date: "2026-02-12"
categories:
  - R
  - Data cleaning
  - Package
draft: true
css: _extensions/html/styles.css
editor_options: 
  chunk_output_type: inline
execute:
  echo: false
  warning: false 
fig-align: center
fig-format: png
---

```{r echo=TRUE}
#| code-fold: false
library(thekidsbiostats) # install with install.packages("thekidsbiostats", repos = "the-kids-biostats.r-universe.dev")
```

# Overview

Datasets rarely come with cleaned, tidy formatting in their column names. It is common practice to rename the original column names to something more syntactic, code-friendly names (i.e., lower-case, no spaces or special characters and unique). Alongside code-friendly columns, it is also often useful to store *label attributes*, which provide clean, human-readable versions labels for each code-friendly variable for use in analytical reports, tables and figures. 

While a variety of functions for *renaming columns* (e.g. `dplyr::rename()`) and *applying labels* (e.g., using the [`labelled`](https://cran.r-project.org/web/packages/labelled/index.html) or [`Hmisc`](https://cran.r-project.org/web/packages/Hmisc/index.html) packages) exist, there is still a reasonable amount of leg work needed before such functions can be applied.

This post will demonstrates how the process of standardising and labelling columns can be streamlined and expedited. Specifically, we will introduce use two main functions from [`thekidsbiostats`](https://github.com/The-Kids-Biostats/thekidsbiostats):

+ [`make_column_dict()`](https://the-kids-biostats.github.io/thekidsbiostats/reference/make_column_dict.html) –-- create a copy-pasteable column dictionary template.
+ [`update_columns()`](https://the-kids-biostats.github.io/thekidsbiostats/reference/update_columns.html) –-- rename and label columns using a column dictionary.

# Motivating Example

As an example, let's point to the `data_patient` dataset, which is conveniently contained in `thekidsbiostats` package:

```{r echo=TRUE}
#| code-fold: false
data(data_patient, package = "thekidsbiostats")
```

Previewing this dataset:

```{r}
data_patient %>%
  thekids_table()
```

Immediately, we see that the column headers are not formatted in a very nice way. The column headers contain (any number of):

+ Categorical variable coding
+ Full questions
+ Spacing
+ Inconsistent letter capitalisation
+ Punctuation (parentheses, equals signs, questions marks, dashes, slashes)

The spaces and punctuation in these column names require us to use backticks (``) to avoid R interpreting these symbols literally. Beyond visually looking messy, it can also be time consuming and tricky to reference these column names when attempting to use them in operations. 

For example, using the original column names, let's implement two trivial manipulations/changes to the data:

::: panel-tabset
## i) Calculate BMI

```{r echo=TRUE}
#| code-fold: false
data_patient$bmi <- data_patient$`WHAT IS YOUR CURRENT WEIGHT? (kg)` / (data_patient$`WHAT IS YOUR HEIGHT? (cm)` / 100.0)^2
```

## ii) Re-Code `Sex`

Recode the Sex column by replacing 0/1 with Male/Female, as defined in the existing column header.

```{r echo=TRUE}
#| code-fold: false
data_patient$`Sex (0=Male, 1=Female)` = case_when(
  data_patient$`Sex (0=Male, 1=Female)` == 0 ~ "Male",
  data_patient$`Sex (0=Male, 1=Female)` == 1 ~ "Female",
  TRUE ~ NA_character_
)
```
:::
::: column-margin
Clearly, coding each of these changes are cumbersome and time intensive, and can restrict the readability of our code workflow.
:::


Functions such as `janitor::clean_names()` can be very useful for standardising such column names into syntactic names. However, this approach is still imperfect, and the resulting names may remain just as problematic (as seen below).

```{r echo=TRUE}
#| code-fold: false
data_patient |>
  select(6) |>
  janitor::clean_names() %>%
  colnames(.)
```
::: column-margin
The "cleaned" version of this column name simply omits spaces and punctuation, and is still far too long.
:::


## Create Data Dictionary

::: callout-important
Renaming columns for easy manipulation in code is potentially dropping any useful formatting of the columns (e.g., units and symbols). 

Ideally, we want to both rename the columns and simultaneously retain their human-readable labels for presentation/reporting.
:::

Our solution to this is therefore to manually set up a dictionary that maps the **old column names** to **newly formatted column names** as well as **apply label attributes**.

The `make_column_dict()` function makes this process slightly easier by creating a ready-to-fill template based on the column names that exist with the data. 

For example:

```{r echo=TRUE}
#| code-fold: false
dict <- make_column_dict(data = data_patient,
                         auto_clean = TRUE)
```

We can either directly add to, or modify, the `dict` object (as assigned above) or copy-paste the printed output into our code and modify the new column names and labels as needed:

```{r echo=TRUE}
#| code-fold: false
dict_patient <- tribble(
  ~old, ~new, ~label,
  'Patient ID#', 'id', 'Patient ID',
  'DOB [YYYY-MM-DD]', 'dob', 'Date of Birth',
  'Sex (0=Male, 1=Female)', 'sex', 'Sex',
  'WHAT IS YOUR HEIGHT? (cm)', 'height_cm', 'Height (cm)',
  'WHAT IS YOUR CURRENT WEIGHT? (kg)', 'weight_kg', 'Weight (kg)',
  'Do you currently smoke any form of tobacco products, including cigarettes, cigars, or pipes, on a regular basis?', 'current_smoking', 'Current tobacco consumption',
  'bp (mmHg)', 'bp_mmhg', 'BP (mmHg)',
  'Cholesterol / mmolL', 'cholesterol_mmol_l', 'Cholesterol (mmolL)',
)
```

Alternatively, we can also write this file (e.g., csv, rds, xlsx) to our machine, by specifying `file`:

```{r echo=TRUE, eval=FALSE}
#| code-fold: false
make_column_dict(data = data_patient,
                 auto_clean = TRUE,
                 file = "colnames.csv")
```

And then read the csv back in after new column names and labels have been added:

```{r echo=TRUE}
#| code-fold: false
dict_patient <- readr::read_csv(file = "colnames.csv")
```

```{r}
dict_patient %>%
  thekids_table()
```

## Apply column dictionary

Now that we have defined the column dictionary above, we can simply use the `update_columns()` function to apply these changes in a single line.

```{r echo=TRUE}
#| code-fold: false
data_clean <- update_columns(data_patient, dict = dict_patient)
```
::: column-margin
The "default" names for the old, new and label attributes bust
:::

Previewing this "cleaned" dataset:

```{r}
data_clean %>%
  labelled::remove_labels() %>%
  thekids_table()
```

Because the labels we specified are stored as attributes, they are automatically recognised by many summary packages that respect label metadata (e.g., `gtsummary::tbl_summary()` tables).

```{r}
data_clean %>%
  tbl_summary(include = -id,
              type = list(height_cm ~ "continuous",
                          weight_kg ~ "continuous",
                          cholesterol_mmoll ~ "continuous",
                          bmi ~ "continuous")) %>%
  thekids_table()
```


# Closing comments

Overall, the `make_column_dict()` and `update_columns()` functions from our package offer an efficient solution for transforming and labelling datasets with problematic column headers. More concise column names are easier to work with and enhance the readability/tractability of code, while user-defined labels allow users to easily produce more polished and publication-ready tables and figures.

# Acknowledgements

Thanks to Dr Matt Cooper, Wesley Billingham, Dr Robin Cook and Dr Elizabeth McKinnon for providing feedback on and reviewing this post.

## Reproducibility Information

To access the .qmd (Quarto markdown) files as well as any R scripts or data that was used in this post, please visit our GitHub:

<https://github.com/The-Kids-Biostats/The-Kids-Biostats.github.io/tree/main/posts/>

The session information can also be seen below.

```{r echo = T}
sessionInfo()
```
