---
title: "An Overview of the 'thekidsbiostats' R package"
author: "Wesley Billingham"
format:
  html:
    code-fold: true
    tabset: true
    toc: true
    toc-location: left
date: "2025-01-15"
categories:
  - R
  - Summary
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
suppressMessages(library(Hmisc))
#library(gtsummary)
#library(summarytools)
#library(desctable)
#library(tableone)
```

# Overview

A major attraction of the R programming language is its vast library of free packages, which can be found on both CRAN (the Comprehensive R Archive Network) and hosted on other sites such as GitHub. These packages can provide ready-made functions, datasets and/or templates for all kinds of applications (some staggeringly niche). In the field of statistics and data analysis, it is rare to encounter any methodology, model and/or algorithm which has not also been implemented in an R package somewhere.

Packages can be easily developed for use by individuals or teams, to automate or simplify common tasks. This is the purpose of `thekidsbiostats` for our teams. Its two primary purposes can be broadly described as providing:

  1. Functions to automate and or simplify regular tasks in our workflow. 
  2. Functions and Quarto templates which apply consistent formatting and theming to documents, ggplots, tables, etc. 
  
This post will provide short explanations and examples of our most frequently used functions. While some of the formatting functions and templates may seem at first glance to be specific to The Kids Research Institute Australia (TKRIA), they can be readily adjusted for use in other contexts with changes to fonts, colours, etc. 

## Installing the Package

The package (and all of its code!) are available here:
https://github.com/The-Kids-Biostats/thekidsbiostats

To install and load the package, simply run the code below:

```{r echo = T}
remotes::install_github("https://github.com/The-Kids-Biostats/thekidsbiostats")

library(thekidsbiostats)
```

## Regular Tasks

The following five functions exist to simplify repetitive tasks. Whether it is worth the time to automate such a task is helpfully informed by this xkcd:

### create_project



### create_template

### thekids_model

### round_vec and round_df

These two functions are examples where the sole purpose is keep code concise and to the point. Often when we present figures whether it is in plots, tables or in written form, we wish to keep trailing zeroes when rounding numbers to a x decimal places. This is not the default behaviour of the round() function in R. Below we demonstrate the difference between round() and our round_vec() function.

```{r echo = F}
data.frame(
  original = c(1.8003, 1.9998, 2.5812)
) %>%
  mutate(
    after_round = round(original, 2),
    after_round_vec = round_vec(original, 2)
  ) %>%
  kable() %>% kable_styling()
  thekids_table()

```




### fct_case_when



## Theming

### thekids_theme

### thekids_table


# Conclusions

Overall, the `gtsummary::tbl_summary` is the best summary table function out of those explored in this blog. In general, the range of in-built parameters and helper functions allows for the greatest flexibility in producing a relatively publication-ready summary table.

## Acknowledgements

Thanks to Wesley Billingham, Matt Cooper, and Elizabeth McKinnon for providing feedback on and reviewing this post.

You can look forward to seeing posts from these other team members here in the coming weeks and months.

## Reproducibility Information

To access the .qmd (Quarto markdown) files as well as any R scripts or data that was used in this post, please visit our GitHub:

<https://github.com/The-Kids-Biostats/The-Kids-Biostats.github.io/tree/main/posts/>

The session information can also be seen below.

```{r echo = T}
sessionInfo()
```

