---
title: "Interaction Terms in Logistic Regression"
author: "Zac Dempsey"
format:
  html:
    code-fold: true
    toc: true
    toc-location: left
date: Sys.Date()
categories:
  - R
  - regression
  - model
  - interpretation
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

-   Anybody can fit a model to any data set. And R will oblige so long as it is possible!

-   Interpreting/drawing meaning from model coefficients is where the bulk of the work/complexity arises in statistical/biostatistical work. Also what we need to write to deliver at the end of the day!

    -   Interaction terms add an additional layer of complexity, particularly with respect to nonlinear models.

# Background

## Logistic Regression

### Components

-   Odds

-   Log-odds

-   Proportions

### Model Form

-   Coefficients are estimated on a log-odds scale. For interpretation, we often interpret in terms of odds ratio (exponentiation)

## Interaction Terms

# Motivating Example

-   Leave the correct interpretation hanging? Can we structure things to set up a motivating example

# Example 1

## Acknowledgements

Thanks to Wesley Billingham, Matt Cooper, and Elizabeth McKinnon for providing feedback on and reviewing this post.

You can look forward to seeing posts from these other team members here in the coming weeks and months.

<br>

## Reproducibility Information

To access the .qmd (Quarto markdown) files as well as any R scripts or data that was used in this post, please visit our GitHub:

The session information can also be seen below.

```{r echo = T}
sessionInfo()
```
