---
title: "Summary Tabling in R"
author: "Zac Dempsey"
format:
  html:
    code-fold: true
    toc: true
    toc-location: left
date: "2024-07-01"
categories:
  - R
  - Table
  - Summary
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```         
```

<br>

## Acknowledgements

Thanks to Wesley Billingham, Matt Cooper, and Elizabeth McKinnon for providing feedback on and reviewing this post.

You can look forward to seeing posts from these other team members here in the coming weeks and months.

<br>

## Reproducibility Information

To access the .qmd (Quarto markdown) files as well as any R scripts or data that was used in this post, please visit our GitHub:

<https://github.com/The-Kids-Biostats/The-Kids-Biostats.github.io/tree/main/posts/parallel>

The session information can also be seen below.

```{r echo = T}
sessionInfo()
```

```{r echo = F , warning = F, message = F, error = F}
rm(trials, results, i, j, model, outcome_vars, input_vars, start, end, ind, bootstrap_results, bootstrap_coefs, bootstrap_cis, mtcars, result, month_dat, flights)
```
