---
title: "Efficiently Cleaning and Labelling Data Columns with `thekidsbiostats`"
author: "Zac Dempsey, Dr Robin van de Meeberg"
format:
  html:
    code-fold: true
    tabset: true
    toc: true
    toc-location: left
image: "favicon.svg"
date: "2026-02-19"
categories:
  - R
  - Data cleaning
  - Package
draft: false
css: _extensions/html/styles.css
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false 
fig-align: center
fig-format: png
---

```{r echo=TRUE}
#| code-fold: false
library(thekidsbiostats) # install with install.packages("thekidsbiostats", repos = "the-kids-biostats.r-universe.dev")
```

# Overview

Datasets rarely come with clean and consistent formatting in their column names. Therefore, it is common practice when performing analysis to rename the original column names to coding-friendly names (i.e., unique, lower-case, no spaces or special characters, consistent prefixes/suffixes).

Alongside these coding-friendly columns, it is then also useful to store (or link) *label attributes*, which provide clean, more descriptive or verbose labels for each coding-friendly variable name. These can be called upon when we are producing output for presentation, such as complete analytical reports, or exported tables and figures.

While a variety of functions for **renaming columns** (e.g., `dplyr::rename()`) and **applying labels** (e.g., using the [`labelled`](https://cran.r-project.org/web/packages/labelled/index.html) or [`Hmisc`](https://cran.r-project.org/web/packages/Hmisc/index.html) packages) exist, a reasonable amount of leg work is needed to use such functions across an entire dataset.

This post will demonstrate how two new functions within [`thekidsbiostats`](https://github.com/The-Kids-Biostats/thekidsbiostats) package can be used to streamline and expedite the process of standardising and labelling columns. Specifically, let's introduce the following functions:

-   [`make_column_dict()`](https://the-kids-biostats.github.io/thekidsbiostats/reference/make_column_dict.html) — create a copy-pasteable column name dictionary template.
-   [`update_columns()`](https://the-kids-biostats.github.io/thekidsbiostats/reference/update_columns.html) — renames and labels columns using a column name dictionary.

# Motivating Example

As an example, let's explore the `data_patient` dataset, an example dataset shipped with the package:

```{r echo=TRUE}
#| code-fold: false
data(data_patient, package = "thekidsbiostats")
```

Previewing this dataset:

```{r}
data_patient %>%
  thekids_table() |> 
  bg(bg=thekids_palettes$tint10[['CoolGrey']]) |> 
  modify_labels(hline=TRUE)
```

Immediately, we see that the column headers are not formatted in a very nice way, and especially not for coding purposes! The column headers contain (any combination of):

-   Categorical variable coding,
-   Full questions,
-   Spacing,
-   Inconsistent letter capitalisation,
-   Punctuation (parentheses, equals signs, questions marks, dashes, slashes).

The spaces and punctuation in these column names require us to use backticks (\`\`) to avoid R interpreting these symbols literally.

Beyond visually looking messy within final output, this set up can also be messy within your code, unnecessarily making simple operations more difficult to follow. For example, when forced to use the original column names, let's see what implementing two trivial manipulations/changes to the data looks like in the code:

::: panel-tabset
## i) Calculate BMI

```{r echo=TRUE}
#| code-fold: false
data_patient$bmi <- data_patient$`WHAT IS YOUR CURRENT WEIGHT? (kg)` / (data_patient$`WHAT IS YOUR HEIGHT? (cm)` / 100.0)^2
```

## ii) Re-code Sex variable

Recode the Sex column by replacing 0/1 with Male/Female, as defined in the existing column header.

```{r echo=TRUE}
#| code-fold: false
data_patient$`Sex (0=Male, 1=Female)` = case_when(
  data_patient$`Sex (0=Male, 1=Female)` == 0 ~ "Male",
  data_patient$`Sex (0=Male, 1=Female)` == 1 ~ "Female",
  TRUE ~ NA_character_
)
```
:::

::: column-margin
Clearly, coding each of these changes is more cumbersome and time intensive than it needs to be, and the readability of our code is impacted.
:::

<br>

Functions such as `janitor::clean_names()` can be very useful for standardising such column names into coding-friendly names. However, this approach is still imperfect, and the resulting names may remain just as problematic (as seen below).

```{r echo=TRUE}
#| code-fold: false
colnames(janitor::clean_names(data_patient[6]))
```

::: column-margin
The "cleaned" version of this column name simply omits spaces and punctuation, however, it is still far too long.
:::

# Our Solution

Ideally, we want to simultaneously

1.  rename the columns, and
2.  retain human-readable labels for presentation/reporting.

Therefore, our solution is to set up a dictionary that maps the **old column names** to **newly-formatted column names** as well as **apply label attributes**.

## Step 1: Create Data Dictionary

The `make_column_dict()` function facilitates this process by giving you a ready-to-fill template based on the original column names in the data.

For example:

```{r echo=TRUE}
#| code-fold: false
dict <- make_column_dict(data = data_patient, auto_clean = TRUE, quiet = FALSE)
```

::: column-margin
The output of `make_column_dict` is a `tribble` — a *row-wise* data-frame object (or `tibble`) which allows one to more easily visualise the format of the table.
:::

We can either directly add to, or modify, the `dict` object assigned above or copy-paste the output into our code and add/edit/update column names and labels as needed (see below):

```{r echo=TRUE}
#| code-fold: false
dict_patient <- tribble(
  ~old, ~new, ~label,
  'Patient ID#', 'id', 'Patient ID',
  'DOB [YYYY-MM-DD]', 'dob', 'Date of Birth',
  'Sex (0=Male, 1=Female)', 'sex', 'Sex',
  'WHAT IS YOUR HEIGHT? (cm)', 'height_cm', 'Height (cm)',
  'WHAT IS YOUR CURRENT WEIGHT? (kg)', 'weight_kg', 'Weight (kg)',
  'Do you currently smoke any form of tobacco products, including cigarettes, cigars, or pipes, on a regular basis?', 'current_smoking', 'Currently consumes tobacco',
  'bp (mmHg)', 'bp_mmhg', 'BP (mmHg)',
  'Cholesterol / mmolL', 'cholesterol_mmol_l', 'Cholesterol (mmolL)',
  'bmi', 'bmi', "BMI (kg/m^2)"
)
```

::: column-margin
This `tribble` has been manually modified with "clean" column names and variable labels. Additionally, a row for the derived `bmi` variable has been included.
:::

::: callout-important
Naively renaming columns for easy manipulation in code risks dropping potentially useful information contained in the original column name (e.g., units and symbols). It is important to have a *data dictionary* to retain this information about variables, which is distinct from the column name dictionary we're creating here. For example, neither our `new` or `label` value for the `sex` variable above captures the coding.
:::

As the number of columns in your dataset increases, editing/maintaining this within your code may become increasingly unwieldy. The option is also available to manage this via an external file (e.g., *.csv*, *.rds*, *.xlsx*), by specifying `file`:

```{r echo=TRUE, eval=FALSE}
#| code-fold: false
make_column_dict(data = data_patient,
                 auto_clean = TRUE,
                 file = "colnames.csv")
```

If need be, we could read the dictionary back into R after having modified (in an external program) the column names and labels as necessary:

```{r echo=TRUE}
#| code-fold: false
dict_patient <- readr::read_csv(file = "colnames.csv")
```

```{r}
dict_patient %>%
  thekids_table(padding.left=8, padding.right=6, padding.bottom=0, padding.top=0) |> 
  bg(bg=thekids_palettes$tint10[['CoolGrey']]) |> 
  modify_labels(hline=TRUE)
```

## Step 2: Applying the column dictionary

Now that we have defined the column dictionary above, we can simply use `update_columns()` to apply these changes in a single line.

```{r echo=TRUE}
#| code-fold: false
data_clean <- update_columns(
  data_patient, 
  dict = dict_patient
)
```

::: column-margin
The `update_columns()` function includes arguments `old`, `new`, `label` and these must correspond to the column names of the data dictionary object. By default, the names assigned by `make_column_dict()` are assumed. It is unlikely you will have to modify these arguments.
:::

## Previewing Output

Previewing this "cleaned" dataset:

```{r}
data_clean %>%
  labelled::remove_labels() %>%
  thekids_table() |> 
  bg(bg=thekids_palettes$tint10[['CoolGrey']]) |> 
  modify_labels(hline=TRUE)
```

Because the labels we specified are stored as attributes, they are automatically recognised by many summary packages that respect label metadata (e.g., `gtsummary::tbl_summary()` tables).

<br>

So after the steps above, we no longer have this...

```{r echo=TRUE}
#| code-fold: false
data_patient %>%
  tbl_summary(include = -`Patient ID#`,
              type = list(`WHAT IS YOUR HEIGHT? (cm)` ~ "continuous",
                          `WHAT IS YOUR CURRENT WEIGHT? (kg)` ~ "continuous",
                          `Cholesterol / mmolL` ~ "continuous",
                          `bmi` ~ "continuous")) %>%
  thekids_table(padding=2) |>
  modify_labels(hline=TRUE, bold=TRUE)
```

We have this!

```{r echo=TRUE}
#| code-fold: false
data_clean %>%
  tbl_summary(include = -id,
              type = list(height_cm ~ "continuous",
                          weight_kg ~ "continuous",
                          cholesterol_mmol_l ~ "continuous",
                          bmi ~ "continuous")) %>%
  thekids_table(padding=2) |> 
  modify_labels(hline=TRUE, bold=TRUE)
```

# Closing comments

Overall, the `make_column_dict()` and `update_columns()` functions offer an efficient solution for transforming and labelling datasets with problematic column headers. More concise column names are easier to work with and enhance the readability/tractability of code, while user-defined labels allow users to easily produce more polished and publication-ready outputs.

# Acknowledgements

Thanks to Dr Matt Cooper, Wesley Billingham, Dr Haileab Wolde, and Dr Elizabeth McKinnon for providing feedback on and reviewing this post.

## Reproducibility Information

To access the .qmd (Quarto markdown) files as well as any R scripts or data that was used in this post, please visit our GitHub:

<https://github.com/The-Kids-Biostats/The-Kids-Biostats.github.io/tree/main/posts/>

The session information can also be seen below.

```{r echo = T}
sessionInfo()
```
